paper

### 2024-09-19 paper.qmd
paper work

@@

limrots.result <- readRDS("../LimROTS paper/UPS1.Case0.LimROTS.rds")
rots.result <- readRDS("../LimROTS paper/UPS1.Case0.ROTS.rds")
limma.result <- readRDS("../LimROTS paper/UPS1.Case0.limma.rds")
t.test.result <- readRDS("../LimROTS paper/UPS1.Case0.t.test.rds")

@@

@@

metrics_limrots <- extract_metrics(conf_matrix_limrots, "limROTS")
metrics_rots <- extract_metrics(conf_matrix_rots, "ROTS")
metrics_limma <- extract_metrics(conf_matrix_limma, "limma")
metrics_t.test <- extract_metrics(conf_matrix_t.test, "t.test")

all_metrics <- bind_rows(metrics_limrots, metrics_rots, metrics_limma,metrics_t.test)

@@

-   **limma** performs significantly worse than both **limROTS** and **ROTS**, especially in PPV and F1 score, making it less reliable for this particular dataset and analysis task. The lower sensitivity and specificity suggest that limma struggles with both true positive and true negative identification compared to the other two methods.

@@

@@

limrots.result <- readRDS("../LimROTS paper/UPS1.Case1.LimROTS.rds")
rots.result <- readRDS("../LimROTS paper/UPS1.Case1.ROTS.rds")
limma.result <- readRDS("../LimROTS paper/UPS1.Case1.limma.rds")
ancova.result <- readRDS("../LimROTS paper/UPS1.Case1.ancova.rds")

alpha_thresholds <- seq(0, 0.2, by = 0.001)
significant_limROTS <- sapply(alpha_thresholds, function(alpha) sum(limrots.result$q_values$qvalues < alpha))
significant_ROTS <- sapply(alpha_thresholds, function(alpha) sum(rots.result$FDR < alpha))
significant_limma <- sapply(alpha_thresholds, function(alpha) sum(limma.result$adj.P.Val < alpha))
significant_ancova <- sapply(alpha_thresholds, function(alpha) sum(ancova.result$FDR < alpha))

plot_data <- data.frame(Alpha = alpha_thresholds, significant_limROTS = significant_limROTS,
                        significant_ROTS = significant_ROTS , significant_limma = significant_limma  ,significant_ancova = significant_ancova)

@@

@@

limrots.result <- readRDS("paperdata/BM21.Case0.limrots.rds")
rots.result <- readRDS("paperdata/BM21.Case0.rots.rds")
limma.result <- readRDS("paperdata/BM21.Case0.limma.rds")
t.test.result <- readRDS("paperdata/BM21.Case0.t.test.rds")

alpha_thresholds <- seq(0, 0.075, by = 0.01)
significant_limROTS <- sapply(alpha_thresholds, function(alpha) sum(limrots.result$q_values$qvalues < alpha))
significant_ROTS <- sapply(alpha_thresholds, function(alpha) sum(rots.result$FDR < alpha))
significant_limma <- sapply(alpha_thresholds, function(alpha) sum(limma.result$adj.P.Val < alpha))
significant_t.test <- sapply(alpha_thresholds, function(alpha) sum(t.test.result$fdr < alpha))

plot_data <- data.frame(Alpha = alpha_thresholds, significant_limROTS = significant_limROTS,
                        significant_ROTS = significant_ROTS , significant_limma = significant_limma, significant_t.test = significant_t.test )

@@

# Run LimROTS analysis with trend and robust settings enabled
limrots.result <- LimROTS(data.exp = BM21.Case1, 
                          B = B, K = K, meta.info = meta.info, 
                          cluster = cluster, group.name = group.name, 
                          formula.str = formula.str, trend = TRUE, robust = TRUE, n.ROTS = TRUE)

saveRDS(limrots.result , "paperdata/BM21.Case1.rots.rds")
```

@@

data(BM21.Case1)
print(BM21.Case1)
data <- assay(BM21.Case1)
formula.str <- "~ 0 + mixing.ratio"
meta.info <- data.frame(colData(BM21.Case1), check.names = FALSE, row.names = row.names(colData(BM21.Case1)))
meta.info$sample.id <- NULL
design.matrix <- model.matrix(formula(formula.str), data = meta.info)
colnames(design.matrix) <- make.names(colnames(design.matrix))
fit <- lmFit(data, design.matrix)
cont_matrix <- makeContrasts("mixing.ratio1-mixing.ratio2",  levels=design.matrix)
fit2 <- contrasts.fit(fit, cont_matrix)
fit.ebayes <- eBayes(fit2, trend=T, robust=T)
limma.result <- topTable(fit.ebayes, coef="mixing.ratio1-mixing.ratio2" , number = "Inf")
```

@@

# Load the dataset
load("../LimROTS paper/metabolomics/BM21.Case3.rda")

print(BM21.Case3)

@@

# Set metadata and formula for LimROTS analysis
meta.info <- c("mixing.ratio" , "fake.batch")
B <- 1000  # Number of bootstrap samples
K <- dim(BM21.Case3)[1]/4  # Set the value for K based on the data size
K <- floor(K) 
num_cores <- 30   # Number of cores for parallel processing
cluster <- makeCluster(num_cores)  # Create a parallel cluster
group.name <- "mixing.ratio"
formula.str <- "~ 0 + mixing.ratio + fake.batch"  # Formula for group comparison

@@

@@

length(which(limma.result$adj.P.Val < 0.05))

limma.result <- limma.result[limma.result$adj.P.Val < 0.05,]

>>>>

### Run samr

```{r}
library(samr)
data("UPS1.Case0")
data <- assay(UPS1.Case0)
meta.info <- data.frame(colData(UPS1.Case0), check.names = FALSE, row.names = row.names(colData(UPS1.Case0)))

data <- list(x = as.matrix(data), y = rev(meta.info$Conc.), geneid = row.names(data), genenames = row.names(data), logged2 = TRUE)

# Set up SAM input data
samfit <- samr(data, resp.type="Two class unpaired", nperms = 1000 )

delta.table <- samr.compute.delta.table(samfit, min.foldchange = 0)

delta <- delta.table[delta.table[, "median FDR"] < 0.05, ][1, ][1] 
siggenes.table <- samr.compute.siggenes.table(samfit, delta, data, delta.table, min.foldchange = 0 , all.genes = TRUE)

both <- rbind(siggenes.table$genes.up, siggenes.table$genes.lo)
dim(both)
both <- data.frame(both)

both$q.value... <- as.numeric(both$q.value...) / 100

saveRDS(both,"paperdata/UPS1.Case0.samr.rds")
```

@@

data(UPS1.Case0)

limrots.result <- readRDS("paperdata/UPS1.Case0.LimROTS.rds")
rots.result <- readRDS("paperdata/UPS1.Case0.ROTS.rds")
limma.result <- readRDS("paperdata/UPS1.Case0.limma.rds")
t.test.result <- readRDS("paperdata/UPS1.Case0.t.test.rds")
samr.result <- readRDS("paperdata/UPS1.Case0.samr.rds")

@@

predictions_samr <- samr.result$q.value... < 0.05
predictions_samr <- factor(predictions_samr, levels = c(TRUE, FALSE))
true_labels_samr <- ifelse(samr.result$Gene.ID %in% TP , TRUE, FALSE)
true_labels_samr <- factor(true_labels_samr, levels = c(TRUE, FALSE))
conf_matrix_samr <- confusionMatrix(predictions_samr, true_labels_samr)

@@

metrics_limrots <- extract_metrics(conf_matrix_limrots, "limROTS")
metrics_rots <- extract_metrics(conf_matrix_rots, "ROTS")
metrics_limma <- extract_metrics(conf_matrix_limma, "limma")
metrics_t.test <- extract_metrics(conf_matrix_t.test, "t.test")
metrics_samr <- extract_metrics(conf_matrix_samr, "samr")

all_metrics <- bind_rows(metrics_limrots, metrics_rots, metrics_limma,metrics_t.test , metrics_samr)

@@

-   **limma, and samr** performs significantly worse than both **limROTS** and **ROTS**, especially in PPV and F1 score, making it less reliable for this particular dataset and analysis task. The lower sensitivity and specificity suggest that limma struggles with both true positive and true negative identification compared to the other two methods.

@@

### Run samr

```{r}
library(samr)

data("UPS1.Case1")
data <- assay(UPS1.Case1)
meta.info <- data.frame(colData(UPS1.Case1), check.names = FALSE, row.names = row.names(colData(UPS1.Case1)))

data <- list(x = as.matrix(data), y = rev(meta.info$Conc.), geneid = row.names(data), genenames = row.names(data), logged2 = TRUE)

# Set up SAM input data
samfit <- samr(data, resp.type="Two class unpaired", nperms = 1000 )

delta.table <- samr.compute.delta.table(samfit, min.foldchange = 0)

delta <- delta.table[delta.table[, "median FDR"] < 0.05, ][1, ][1] 
siggenes.table <- samr.compute.siggenes.table(samfit, delta, data, delta.table, min.foldchange = 0 , all.genes = TRUE)

both <- rbind(siggenes.table$genes.up, siggenes.table$genes.lo)
dim(both)
both <- data.frame(both)

both$q.value... <- as.numeric(both$q.value...) / 100

saveRDS(both,"paperdata/UPS1.Case1.samr.rds")
```

@@

limrots.result <- readRDS("paperdata/UPS1.Case1.LimROTS.rds")
rots.result <- readRDS("paperdata/UPS1.Case1.ROTS.rds")
limma.result <- readRDS("paperdata/UPS1.Case1.limma.rds")
ancova.result <- readRDS("paperdata/UPS1.Case1.ancova.rds")
samar.result <- readRDS("paperdata/UPS1.Case1.samr.rds")

alpha_thresholds <- seq(0, 0.2, by = 0.001)
significant_limROTS <- sapply(alpha_thresholds, function(alpha) sum(limrots.result$q_values$qvalues < alpha))
significant_ROTS <- sapply(alpha_thresholds, function(alpha) sum(rots.result$FDR < alpha))
significant_limma <- sapply(alpha_thresholds, function(alpha) sum(limma.result$adj.P.Val < alpha))
significant_ancova <- sapply(alpha_thresholds, function(alpha) sum(ancova.result$FDR < alpha))
significant_samr <- sapply(alpha_thresholds, function(alpha) sum(samar.result$q.value... < alpha))

plot_data <- data.frame(Alpha = alpha_thresholds, significant_limROTS = significant_limROTS,
                        significant_ROTS = significant_ROTS , significant_limma = significant_limma  ,significant_ancova = significant_ancova , significant_samr = significant_samr)

@@

### Run samar

```{r}
library(samr)

data("BM21.Case0")
data <- assay(BM21.Case0)
meta.info <- data.frame(colData(BM21.Case0), check.names = FALSE, row.names = row.names(colData(BM21.Case0)))

data <- list(x = as.matrix(data), y = rev(meta.info$mixing.ratio), geneid = row.names(data), genenames = row.names(data), logged2 = TRUE)

# Set up SAM input data
samfit <- samr(data, resp.type="Two class unpaired", nperms = 1000 )

delta.table <- samr.compute.delta.table(samfit, min.foldchange = 0)

delta <- delta.table[delta.table[, "median FDR"] < 0.05, ][1, ][1] 
siggenes.table <- samr.compute.siggenes.table(samfit, delta, data, delta.table, min.foldchange = 0 , all.genes = TRUE)

both <- rbind(siggenes.table$genes.up, siggenes.table$genes.lo)
dim(both)
both <- data.frame(both)

both$q.value... <- as.numeric(both$q.value...) / 100

saveRDS(both,"paperdata/BM21.Case0.samr.rds")
```

@@

limrots.result <- readRDS("paperdata/BM21.Case0.limrots.rds")
rots.result <- readRDS("paperdata/BM21.Case0.rots.rds")
limma.result <- readRDS("paperdata/BM21.Case0.limma.rds")
t.test.result <- readRDS("paperdata/BM21.Case0.t.test.rds")
samr.result <- readRDS("paperdata/BM21.Case0.samr.rds")

alpha_thresholds <- seq(0, 0.075, by = 0.01)
significant_limROTS <- sapply(alpha_thresholds, function(alpha) sum(limrots.result$q_values$qvalues < alpha))
significant_ROTS <- sapply(alpha_thresholds, function(alpha) sum(rots.result$FDR < alpha))
significant_limma <- sapply(alpha_thresholds, function(alpha) sum(limma.result$adj.P.Val < alpha))
significant_t.test <- sapply(alpha_thresholds, function(alpha) sum(t.test.result$fdr < alpha))
significant_samr <- sapply(alpha_thresholds, function(alpha) sum(samr.result$q.value... < alpha))

plot_data <- data.frame(Alpha = alpha_thresholds, significant_samr = significant_samr,
                        significant_limROTS = significant_limROTS,
                        significant_ROTS = significant_ROTS , significant_limma = significant_limma, significant_t.test = significant_t.test  )

@@

# Run LimROTS analysis with trend and robust settings enabled
rots.result <- LimROTS(data.exp = BM21.Case1, 
                          B = B, K = K, meta.info = meta.info, 
                          cluster = cluster, group.name = group.name, 
                          formula.str = formula.str, trend = TRUE, robust = TRUE, n.ROTS = TRUE)

saveRDS(rots.result , "paperdata/BM21.Case1.rots.rds")
```

@@

data(BM21.Case1)
print(BM21.Case1)
data <- assay(BM21.Case1)
formula.str <- "~ 0 + mixing.ratio"
meta.info <- data.frame(colData(BM21.Case1), check.names = FALSE, row.names = row.names(colData(BM21.Case1)))
meta.info$sample.id <- NULL
design.matrix <- model.matrix(formula(formula.str), data = meta.info)
colnames(design.matrix) <- make.names(colnames(design.matrix))
fit <- lmFit(data, design.matrix)
cont_matrix <- makeContrasts("mixing.ratio1-mixing.ratio2",  levels=design.matrix)
fit2 <- contrasts.fit(fit, cont_matrix)
fit.ebayes <- eBayes(fit2, trend=T, robust=T)
limma.result <- topTable(fit.ebayes, coef="mixing.ratio1-mixing.ratio2" , number = "Inf")

saveRDS(limma.result , "paperdata/BM21.Case1.limma.rds")
```

### Run ANOVA

```{r}
formula.str <- "~ 0 + mixing.ratio"
data(BM21.Case1)

data <- assay(BM21.Case1)
meta.info <- data.frame(colData(BM21.Case1), check.names = FALSE, row.names = row.names(colData(BM21.Case1)))
meta.info$sample.id <- NULL
ancova.result <- data.frame()
for(i in 1:nrow(data)){
  df.temp <- data[i,]

  data.temp <- data.frame(response = as.numeric(df.temp) , mixing.ratio = meta.info$mixing.ratio)

  ancova_model <- lm(formula(paste("response", formula.str)), data = data.temp)
  coef_table <- anova(ancova_model)
  coef_table <- data.frame(coef_table)
  significant_groups <- coef_table[rownames(coef_table) == "mixing.ratio" , ]
  ancova.result <- rbind(ancova.result, data.frame(protein = row.names(data)[i], p.value =significant_groups[, 5] ))
}
ancova.result$FDR <- p.adjust(ancova.result$p.value, method = "fdr")
saveRDS(ancova.result, "paperdata/BM21.Case1.anova.rds")
```

Run samr

```{r}
library(samr)

data("BM21.Case1")
data <- assay(BM21.Case1)
meta.info <- data.frame(colData(BM21.Case1), check.names = FALSE, row.names = row.names(colData(BM21.Case1)))

data <- list(x = as.matrix(data), y = rev(meta.info$mixing.ratio), geneid = row.names(data), genenames = row.names(data), logged2 = TRUE)

# Set up SAM input data
samfit <- samr(data, resp.type="Multiclass", nperms = 1000 )

delta.table <- samr.compute.delta.table(samfit, min.foldchange = 0)

delta <- delta.table[delta.table[, "median FDR"] < 0.05, ][1, ][1] 
siggenes.table <- samr.compute.siggenes.table(samfit, delta, data, delta.table, min.foldchange = 0 , all.genes = TRUE)

both <- rbind(siggenes.table$genes.up, siggenes.table$genes.lo)
dim(both)
both <- data.frame(both)

both$q.value... <- as.numeric(both$q.value...) / 100

saveRDS(both,"paperdata/BM21.Case1.samr.rds")
```

limrots.result <- readRDS("paperdata/BM21.Case1.limrots.rds")
rots.result <- readRDS("paperdata/BM21.Case1.rots.rds")
limma.result <- readRDS("paperdata/BM21.Case1.limma.rds")
anova.result <- readRDS("paperdata/BM21.Case1.anova.rds")
samr.result <- readRDS("paperdata/BM21.Case1.samr.rds")

alpha_thresholds <- seq(0, 0.075, by = 0.01)
significant_limROTS <- sapply(alpha_thresholds, function(alpha) sum(limrots.result$q_values$qvalues < alpha))
significant_ROTS <- sapply(alpha_thresholds, function(alpha) sum(rots.result$FDR < alpha))
significant_limma <- sapply(alpha_thresholds, function(alpha) sum(limma.result$adj.P.Val < alpha))
significant_anova <- sapply(alpha_thresholds, function(alpha) sum(anova.result$FDR < alpha))
significant_samr <- sapply(alpha_thresholds, function(alpha) sum(samr.result$q.value... < alpha))

plot_data <- data.frame(Alpha = alpha_thresholds, significant_limROTS = significant_limROTS,
                        significant_ROTS = significant_ROTS , significant_limma = significant_limma, significant_anova = significant_anova, significant_samr = significant_samr)

plot_data.melt$size <- ifelse(plot_data.melt$Method == "significant_anova", 2.5,1.2)

# Plot using ggplot2
ggplot(plot_data.melt, aes(x = Alpha, y = FDR, color = Method , size =size)) +
  geom_line() +
  theme_minimal() + scale_size_identity()+
  labs(title = "Significant Proteins vs FDR BM21 Case study 4",
       x = "FDR",
       y = "Number of Significant Proteins") +
   geom_vline(xintercept = 0.01, linetype="dotted", 
                color = "red", size=1) + 
    geom_vline(xintercept = 0.05, linetype="dotted", 
                color = "red", size=1)+
    scale_color_npg() 
    theme(
    plot.title = element_text(face = "bold", color = "black"),
    axis.title = element_text(face = "bold", color = "black"),
    axis.text = element_text(face = "bold", color = "black"),
    axis.ticks = element_line(color = "black")
  )
```

@@

# Load the dataset
data(BM21.Case3)
print(BM21.Case3)

@@

# Set metadata and formula for LimROTS analysis
meta.info <- c("mixing.ratio" , "fake.batch")
B <- 1000  # Number of bootstrap samples
K <- dim(BM21.Case3)[1]/2  # Set the value for K based on the data size
K <- floor(K) 
num_cores <- 30   # Number of cores for parallel processing
cluster <- makeCluster(num_cores)  # Create a parallel cluster
group.name <- "mixing.ratio"
formula.str <- "~ 0 + mixing.ratio + fake.batch"  # Formula for group comparison

@@

length(which(limrots.result.df$q.value < 0.01))

limrots.result.df <- limrots.result.df[limrots.result.df$q.value < 0.01,]

length(which(grepl("_pluseffect", row.names(limrots.result.df)) ))

length(which(grepl("synthetic", row.names(limrots.result.df)) ))
```

@@

length(which(limma.result$adj.P.Val < 0.01))

limma.result <- limma.result[limma.result$adj.P.Val < 0.01,]
